Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0

### 2019-06-07 2019_04_22_Reviewer_Comments.Rmd
updated comments, added some sections to the manuscript, new citations

```
5) Line 208, ".... Then genes (or transcripts), then tissues." The "genes" should be "gene IDs by official gene name". The tables derived from "expression" search (e.g. Fig.4A) should be further annotated, e.g. define what are "decile", "log2(PMT+1)", etc. 
```
Line 208 now reads "The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then Hugo Gene Nomeclature Committee (HGNC) genes names (or ENSEMBL transcripts), then tissues" (it is now line 247). We define decile, log2(TPM + 1) in the Figure 4 text (lines 255-258) as follows "Figure 4: Screenshots from eyeIntegration web app. A. Pan-tissue gene expression box plots with accompanying data tables. The data tables display the rank (lower is more highly expressed) of each gene in each sub tissue, the decile of the rank (10 is the highest decile of expression), and the mean gene’s log2(TPM + 1) score for each sub tissue. B. Heatmap visualization."
```
6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
```

We added a short section to the discussion giving some tips on how to help with future updates to EiaD from lines

@@

Compare to GTEx (their counts against our counts)

library(tidyverse)
gtex <- read_tsv('https://storage.googleapis.com/gtex_analysis_v7/rna_seq_data/GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_tpm.gct.gz', skip = 2)
tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')
metadata <- metadata %>% mutate(si = str_extract(sample_attribute, "GTEX\\S+"))
samples <- metadata$si[metadata$si %in% colnames(gtex)]
genes <- tpm$ID[tpm$ID %in% gtex$Description]

long_tpm <- tpm %>% gather(Sample, TPM, -ID)
long_gtex <- gtex %>% select(-Name) %>% gather(Sample, Count, -Description)

long_tpm2 <- left_join(long_tpm, metadata %>% select(sample_accession, si), by = c("Sample" = "sample_accession")) %>% filter(si %in% samples)

both <- left_join(long_tpm2 %>% select(-Sample) %>% rename(Sample = si), long_gtex %>% filter(Sample %in% samples) %>% rename(ID=Description), by = c('ID','Sample'))

cor(both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(TPM), both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(Count), method = 'spearman')
# spearman 0.91
cor(both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(TPM), both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(Count), method = 'pearson')
# pearson 0.72
save(both, file = 'gtex_eyeIntegration_compare.Rdata')

# reports pearson cor of TPM of 0.68 to 0.72 between alignment free (salmon/kallisto) methods compared to alignment-based methods (HISAT2)
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6042521/

Compare retina / RPE / cornea markers?
```
2. It would be helpful to provide some information about the accuracy of the Salmon transcript quantification tool for the reported analyses. For example, a search for Rhodopsin expression in the single cell RNA seq data shows relatively high expression in cone cells, in which rhodopsin is not usually expressed. Is this an artefact of the scRNA-seq method, or a possible read assignment error from the Salmon tool?
```

The single cell expression data is in active development. Right now we are using the counts quantified by the original research groups. As more scRNA-seq datasets come available we will be re-processing the data ourself.

** UPDATE manuscript to make this more explicit. **

@@

** UPDATE manuscript to list these limitations **

@@

The major way we handle non-biological batch effects is through the use of https://github.com/stephaniehicks/qsmooth, which reshapes the distribution of gene expressions values to be consistent among each tissue type (e.g. Retina/RPE/cornea). We demonstrate the effect of this tool in **LAST PAPER**

While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).

In these situations, I prefer to do less and let the data "speak" for themselves. This is why the primary output of eyeIntegration is a boxplot with the individual data points drawn; this allows the user to get a quick sense of the variance and skew of the data across many datasets.

```
4. Further information about the quality control and filtering applied to the RNA-seq data would be helpful to include in the manuscript. For example, how were the parameters for filtered by a mapping rate of 40% per sample and median read count >2 across all genes per sample selected? Also, basic statistics related to read mapping such as min, max, average, 25 quartile, 75 quartile of mapping should be included. 
```
Parameters selected by checking distributions and picking the parameter that removes the left tail. We have added mapping rate as a supplementary figure and edited the text in the methods accordingly.

@@

I wrote a long explanation of how what we are doing is reasonable, but then thought some more and now have come around to your position. While this t-SNE / euclidean distance cutoff approach works on a empirical level, it is a poor idea.

We have implemented the outlier-by-correlation-to-group approach you suggested (Nature Genetics 46: 430, 2014).
```

@@

We have added the power calculation as supplemental figure.

While it would be nice in *principal* to just use the entire GTEx dataset, there are very practical limiations on why this is inconvenient - the size of the full GTEx datasets, raw, is around 1 petabyte of data, which is very slow to move around and do any compute on. Also this amount of storage space brushes against the limits of our compute cluster.

As for the Ratnapriya/Swaroop data, we mention it briefly in our introduction, but we hadn't added the data to eyeIntegration as it became available shortly after our last data pull.

We have surveyed the SRA / GEO again and have added new studies (including the Ratnapriya/Swaroop) project data to eyeIntegration.

As the AMD data they collected are likely to be of interest, we have made an exception to our "normal tissue" policy and have added the different AMD stages as selectable tissue options.

>>>>

```
5) Line 208, ".... Then genes (or transcripts), then tissues." The "genes" should be "gene IDs by official gene name". The tables derived from "expression" search (e.g. Fig.4A) should be further annotated, e.g. define what are "decile", "log2(PMT+1)", etc. 
```

Line 208 now reads "The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then Hugo Gene Nomeclature Committee (HGNC) genes names (or ENSEMBL transcripts), then tissues" (it is now line 247). We define decile, log2(TPM + 1) in the Figure 4 text (lines 255-258) as follows "Figure 4: Screenshots from eyeIntegration web app. A. Pan-tissue gene expression box plots with accompanying data tables. The data tables display the rank (lower is more highly expressed) of each gene in each sub tissue, the decile of the rank (10 is the highest decile of expression), and the mean gene’s log2(TPM + 1) score for each sub tissue. B. Heatmap visualization."
```
6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
```

We added a short section to the discussion giving some tips on how to help with future updates to EiaD from lines 348-351: "If you wish to have your data added to EiaD in the future, we suggest you 1. deposit data into GEO/SRA, 2. use clear, descriptive, consistent, and detailed metadata for each sample, and 3. (optional) contact the corresponding author. Contacting the corresponding author is only necessary if you feel your data should be included in EiaD and was deposited into the SRA before May 8th, 2019."

@@

To address this concern about the accuracy of our process we directly compared the GTEx provided count table against the counts we calculated for the identical GTEx samples. The code below can be run in R to re-produce our finding: which is that we have a pearson (R^2) correlation of 0.89

```
# R code
# need about 64GB of memory!
library(tidyverse)
gtex <- read_tsv('https://storage.googleapis.com/gtex_analysis_v7/rna_seq_data/GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_tpm.gct.gz', skip = 2)
tpm <- read_tsv('/data/mcgaugheyd/datashare/eyeIntegration/2019_gene_TPM_04.tsv.gz')
metadata <- read_tsv('/data/mcgaugheyd/datashare/eyeIntegration/2019_metadata_04.tsv.gz')
metadata <- metadata %>% mutate(si = str_extract(sample_attribute, "GTEX\\S+"))
samples <- metadata$si[metadata$si %in% colnames(gtex)]
genes <- tpm$ID[tpm$ID %in% gtex$Description]
# make data long to join GTEx and EiaD results
long_tpm <- tpm %>% gather(Sample, TPM, -ID)
long_gtex <- gtex %>% select(-Name) %>% gather(Sample, GTEX_TPM, -Description)
# filter EiaD tpm down to just GTEx samples
long_tpm2 <- left_join(long_tpm, metadata %>% select(sample_accession, si) %>% unique(), by = c("Sample" = "sample_accession")) %>% filter(si %in% samples)
# merge EiaD and GTEx
both <- left_join(long_tpm2 %>% select(-Sample) %>% rename(Sample = si), long_gtex %>% filter(Sample %in% samples) %>% rename(ID=Description), by = c('ID','Sample'))
# remove NA
both2 <- both %>% filter(!is.na(TPM), !is.na(GTEX_TPM))
# calculate correlations with log2(TPM + 0.01 pseudocount) as per Zhang et al. 
cor(log2(both2 %>% pull(TPM) + 0.01), log2(both2 %>% pull(GTEX_TPM) + 0.01), method = 'spearman')
# spearman 0.91
cor(log2(both2 %>% pull(TPM) + 0.01), log2(both2 %>% pull(GTEX_TPM) + 0.01), method = 'pearson')
# pearson 0.89
```

Zhang et al. (https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-4002-1) report a pearson cor R^2 with TPM of 0.89 and 0.93 between alignment free (salmon/kallisto) methods compared to alignment-based methods (STAR/REM) [@zhangEvaluationComparisonComputational2017]. As GTEx used an alignment-based approach (and a different reference genome - we use GRCh38, GTEx uses GRCh37) and we length scale our TPM values (GTEx does note) we feel that this is a reasonable result.

We have added a section to our methods and results section discussing this briefly (lines 126-129, 227-233).

```
2. It would be helpful to provide some information about the accuracy of the Salmon transcript quantification tool for the reported analyses. For example, a search for Rhodopsin expression in the single cell RNA seq data shows relatively high expression in cone cells, in which rhodopsin is not usually expressed. Is this an artefact of the scRNA-seq method, or a possible read assignment error from the Salmon tool?
```

The single cell expression section of eyeIntegration is still in active development. Right now we are using the counts quantified by the original research groups. As more scRNA-seq datasets come available we will be re-processing the data ourself. The high rhodopsin expression is most likely because (some) rods are being incorrectly labelled as cones by the original authors.

The methods section for scRNA-seq procesesing reads "We use the processed gene count data directly from each group, as well as their cluster assignments which specify what cell type each individual cell is. The count data is mean averaged to the cell type, age, and gene level for the single cell expression section of eyeIntegration."

@@

We have added a "limitations" section to the resulst section of the manuscript:

"Salmon quantification, while highly performant and accurate, has a higher variance when with lower read depth and with shorter transcripts [@zhangEvaluationComparisonComputational2017]. Extra care should be taken with comparisons with lower counts of samples (cornea, RGC) as smaller sample numbers decrease the confidence in differential expression. We recommend that if you wish to directly compare your RNA-seq sample quantifications with EiaD, you should not directly compare our TPM values with yours as there are many important variables that will differ. Instead run our Snakemake pipeline (https://www.github.com/davemcg/EiaD_build) and add your samples. Finally, we would like to remind any users that RNA-seq methods measure mRNA levels, but the functional unit is the protein; westerns are still the gold standard with which to evaluate expression and localization."

@@

The major way we handle non-biological batch effects is through the use of https://github.com/stephaniehicks/qsmooth, which reshapes the distribution of gene expressions values to be consistent among each tissue type (e.g. Retina/RPE/cornea).

While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some evidence that these types of tools may overinflate confidence in group differences (see https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/) [@nygaardMethodsThatRemove2016].

We have made a significant change to how our TPM values are computed, using limma's batchEffects() tool to correct the TPM values for mapping rate and sub tissue. These two changes clean up the t-SNE plotting for the eye tissues. The major difference we see within a single sub tissue type is adult retina - for some reason we don't understand the Ratnapriya / Swaroop retina AMD samples (both non-AMD and AMD) cluster apart from all of the other adult retina tissue samples we have collected. This may be related to the lower mapping rate (and RIN scores) from this cohort.

We do not use the tranformed TPM values for the limma / voom differential testing. These transformed values are only used the the t-SNE, boxplots, and the heatmap visualizations.

```
4. Further information about the quality control and filtering applied to the RNA-seq data would be helpful to include in the manuscript. For example, how were the parameters for filtered by a mapping rate of 40% per sample and median read count >2 across all genes per sample selected? Also, basic statistics related to read mapping such as min, max, average, 25 quartile, 75 quartile of mapping should be included. 
```
Parameters were selected by checking distributions and picking the parameter that removes the left tail. We have added mapping rate as a supplementary figure and edited the text in the methods accordingly. Line 111-112: "We first removed samples with a Salmon calculated mapping rate less than 40%. This value was selected as being the far left tail of the distribution of mapping rates across samples (Supplemental Figure 2)."

@@

I first wrote a long explanation of how what we are doing is reasonable, but then thought some more and now have come around to your position. While this t-SNE / euclidean distance cutoff approach works on a empirical level, it is a poor idea.

We have implemented the outlier-by-correlation-to-group approach you suggested (Nature Genetics 46: 430, 2014) (lines 119-125):

"To identify outliers we followed an approach similar to a method in Wright et al [@wrightHeritabilityGenomicsGene2014]. Briefly, we first selected the 3000 genes with the highest variance across all samples and then for each sub-tissue type $T$ and each sample $i$ in $T$, we first calculated $r_{i}$, the average correlation between $i$ and all other samples in $T$. Next, we calculated $D_{i}$, where $D_{i} = \frac{(r_{i} - r\bar{\bar{}})}{median(r_{i} - r\bar{\bar{}})}$ and  $r\bar{\bar{}}$ is the grand mean of all $r_{i}$ for $i$ in $T$. We removed samples with $D_{i} <$ -17.5; we determined this threshold by generating a tSNE plot of our samples, and visually identifying outliers in adult retina tissue.  The ($max(D_{i})$) amongst these outliers was -17.58 and from this we chose -17.5 as our outlier threshold."
```

@@

We have added the power calculation curve as supplemental figure.

While it would be nice in principle to just use the entire GTEx dataset, there are very practical limitations on why this is inconvenient - the size of the full GTEx datasets, raw, is around 1 petabyte of data, which is very slow to move around and do any compute on. Also this amount of storage space brushes against the limits of our allocation in the compute cluster. 
As for the Ratnapriya/Swaroop data, we have now added it to EiaD along with a few other smaller studies which were made available after our last EiaD update in January 2019.

### 2019-06-07 manuscript.Rmd
updated comments, added some sections to the manuscript, new citations

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 1,375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples. The 916 eye and 1,375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD. RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date by database version, gene/transcript, 19 eye tissues, and 54 body tissues. As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members. CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

@@

After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')` and `r supTab_cap(name = 'tSNE_table', display = 'cite')`)). For example, in the retina group, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina samples cluster together. Our ability to uniformly cluster data by known biological source independent of study origin demonstrates that our workflow can effectively account for technical variation between studies.

@@

>>>>

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 1375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples. The 916 eye and 1375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD. RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date by database version, gene/transcript, 19 eye tissues, and 54 body tissues. As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members. CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

To calculate pearson correlation ($R^2$) between GTEx-calculated TPM gene values and our GTEx TPM gene values, we downloaded "GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_tpm.gct.gz" and matched against our GTEx TPM values, running the pearson correlation with $log2(TPM + 0.01)$ values as per Zhang et al. with the cor function in R [@zhangEvaluationComparisonComputational2017].

@@

To ensure there are no substantial differences in quantification of the gene TPM, we calculated the $R^2$ between the GTEX-calculated TPM values and our length scaled and transformed TPM values we calculate for our GTEx samples (see methods). We got an $R^2$ of 0.89. Zhang et al. report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a pearson correlation ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017]. As Zhang et al. compared quantification methods with identical gene references (we use Gencode GRCh38 gene models and GTEx uses hg19) and did not scale TPM score differently, our result falls in line with expectations.

After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')` and `r supTab_cap(name = 'tSNE_table', display = 'cite')`). For example, in the retina group, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina samples cluster together. Our ability to uniformly cluster data by known biological source independent of study origin demonstrates that our workflow can effectively account for technical variation between studies.

@@

## Limitations of the RNA-seq quantification in eyeIntegration

Salmon quantification, while highly performant and accurate, has a higher variance when with lower read depth and with shorter transcripts [@zhangEvaluationComparisonComputational2017]. Extra care should be taken with comparisons with lower counts of samples (cornea, RGC) as smaller sample numbers decrease the confidence in differential expression. We recommend that if you wish to directly compare your RNA-seq sample quantifications with EiaD, you should not directly compare our TPM values with yours as there are many important variables that will differ. Instead run our Snakemake pipeline (https://www.github.com/davemcg/EiaD_build) and add your samples. Finally, we would like to remind any users that RNA-seq methods measure mRNA levels, but the functional unit is the protein; westerns are still the gold standard with which to evaluate expression and localization.
